---
title: "GroupProject"
output:document
date:
---


```{r}
# Install packages
install.packages('skimr')
install.packages('ggplot2')
install.packages("imputeTS")
install.packages("pheatmap")
install.packages("tidyr")
```

```{r}
# Load library
library('imputeTS')
library('skimr')
library('readxl')
library('dplyr')
library('ggplot2')
library('dplyr')
library('pheatmap')
library('tidyr')
library('RColorBrewer')
```
Data preparation:

Exploration of the datasets 
```{r}

# Load datasets
abundances <- read_excel("abundances.xlsx") # Load the abundances dataset
metadata <- read.csv("metadata.csv")       # Load the metadata dataset

# Check the structure of the datasets
str(abundances)  
str(metadata)   

# Summarize the datasets
summary(abundances)  
summary(metadata)    

# Skim the datasets
skim(abundances)  
skim(metadata)    

# Print a preview of the datasets
head(abundances)  
head(metadata)    

```
Filtering of rows with missing values
```{r} 

# Replace string "NA" with actual NA values across all columns
data <- abundances %>%
  mutate(across(everything(), ~ na_if(., "NA")))

# Count the number of NAs in each row
na_counts <- rowSums(is.na(data))

# Define a threshold for filtering rows (e.g., 40% missing data is allowed)
thresholds <- ncol(data) * 0.4

# Identify rows with too many NAs (more than the threshold)
rows_to_filter <- which(na_counts > thresholds)

# Save the list of rows to be filtered out for reference
filtered_out_rows <- data[rows_to_filter, ]

# Print the rows that will be removed
cat("Rows to be filtered out due to excessive missing values:\n")
print(filtered_out_rows)

# Filter rows with missing values exceeding the threshold
abundances_NArowsFiltered <- data %>%
  filter(na_counts <= thresholds)

# Compare row counts before and after filtering
cat("Number of rows before filtering:", nrow(data), "\n")
cat("Number of rows after filtering:", nrow(abundances_NArowsFiltered), "\n")

```
Residual removal of missing values: Change NA to the lowest value of that row
```{r}

# Convert all columns (except the first three) to numeric
abundances_NArowsFiltered <- abundances_NArowsFiltered %>%
  mutate(across(-(1:3), as.numeric))

# Replace NAs with the minimum value in each row (excluding the first three columns)
# Use rowwise() to calculate the minimum for each row and replace NAs
abundance_FilteringDone <- abundances_NArowsFiltered %>%
  rowwise() %>%
  mutate(across(-(1:3), ~ ifelse(is.na(.), min(c_across(-(1:3)), na.rm = TRUE), .))) %>%
  ungroup()

# Check if any NAs are left in the dataset
if (any(is.na(abundance_FilteringDone))) {
  cat("There are still NAs left in the dataset.\n")
} else {
  cat("There are no NAs left in the dataset.\n")
}

# Count remaining NAs (if any)
na_count <- sum(is.na(abundance_FilteringDone))
cat("Total NAs remaining in the dataset:", na_count, "\n")

# Identify rows with NAs
rows_with_na <- which(apply(abundance_FilteringDone, 1, function(x) any(is.na(x))))

# View rows with NAs (if any exist)
if (length(rows_with_na) > 0) {
  cat("Rows with NAs:\n", paste(rows_with_na, collapse = ", "), "\n")
  print(abundance_FilteringDone[rows_with_na, ])  # Print rows containing NAs
} else {
  cat("No rows with NAs found.\n")
}

# Preview the updated dataset
head(abundance_FilteringDone)

```
Removal of excessive columns + changing column names
```{r}
# Remove unnecessary columns ("Symbol", "Protein.names") from the dataset
abundance_FilteringDone_clean <- abundance_FilteringDone %>%
  select(-c(Symbol, Protein.names))

# Rename the first column to "Protein.ID"
colnames(abundance_FilteringDone_clean)[1] <- "Protein.ID"

# Remove non-tumor tissue columns (columns ending with "P")
# Identify and exclude columns with names ending in "P"
abundance_FilteringDone_clean <- abundance_FilteringDone_clean %>%
  select(-matches("P$"))

# 4. Remove the "T" suffix from remaining column names (tumor samples)
colnames(abundance_FilteringDone_clean) <- gsub("T$", "", colnames(abundance_FilteringDone_clean))

# 5. Preview the modified dataset
cat("Updated column names in the cleaned dataset:\n")
print(colnames(abundance_FilteringDone_clean))
head(abundance_FilteringDone_clean)

```
Log transformation
```{r}
# Extract only the abundance data (exclude the first column "Protein.ID")
abundance_only <- as.numeric(unlist(abundance_FilteringDone_clean[, -1]))

# 2. Apply a log transformation (log1p: log(1 + x)) to the abundance data
log_abundance_only <- log1p(abundance_only)

# 3. Preview the transformed data
head(log_abundance_only)

```
Histogram, Q-Q Plot & Density plot: check for normality post log transformation
```{r}
# Custom color palette
custom_colors <- c("pink", "lightblue", "lavender", "lightgreen", "lightcoral", 
                   "lightyellow", "skyblue", "plum", "peachpuff", "thistle")

# Histogram of log-transformed abundance data
ggplot(data.frame(Abundance = log_abundance_only), aes(x = Abundance)) +
  geom_histogram(binwidth = 0.5, fill = custom_colors[1], color = "white", alpha = 0.7) +  
  labs(title = "Histogram of Log-Transformed Proteomic Abundance",
       x = "Log-Transformed Abundance",
       y = "Frequency") +
  theme_minimal(base_size = 14) +  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),  
    axis.title = element_text(face = "bold")  
  )

# Q-Q Plot of log transformed abundance data
qqnorm(log_abundance_only)
qqline(log_abundance_only, col = "red", lwd = 2) 

# Density plot of log transformed abundance data
ggplot(data.frame(Abundance = log_abundance_only), aes(x = Abundance)) +
  geom_density(fill = "pink", alpha = 0.5) +
  labs(title = "Log-transformed Density Plot",
       x = "Log-transformed Proteomic Abundance",
       y = "Density") +
  theme_minimal()

```
Standardisation
```{r}
# Standardize the log-transformed abundance data
# The scale() function standardizes each column (subtracts the mean and divides by the standard deviation)
stand_log_abundance_only <- scale(log_abundance_only)

# View the first few rows of the standardized data
head(stand_log_abundance_only)

# Check if the standardization was performed correctly
# The mean of each column should be approximately 0, and the standard deviation should be approximately 1

# Calculate the mean of each column
column_means <- apply(stand_log_abundance_only, 2, mean)  
cat("Column Means (should be close to 0):\n")
print(column_means)

# Calculate the standard deviation of each column
column_sds <- apply(stand_log_abundance_only, 2, sd)  
cat("Column Standard Deviations (should be close to 1):\n")
print(column_sds)

```     
Heatmap before and after standardisation
```{r}
# Remove the first column if it is non-numeric
# We are assuming the first column is non-numeric (e.g., an identifier)
abundance_FilteringDone_clean_numeric <- abundance_FilteringDone_clean[, -1]

# Check the structure of the dataset to confirm it contains only numeric values
str(abundance_FilteringDone_clean_numeric)

# Convert the numeric dataset to a matrix
abundance_FilteringDone_clean_matrix <- matrix(
  as.numeric(as.matrix(abundance_FilteringDone_clean_numeric)),  # Ensure the data is numeric
  nrow = nrow(abundance_FilteringDone_clean_numeric),
  ncol = ncol(abundance_FilteringDone_clean_numeric)
)

# Optionally convert the matrix to a data frame for further processing
abundance_FilteringDone_clean_df <- as.data.frame(abundance_FilteringDone_clean_matrix)

# Check the structure of the cleaned dataset
str(abundance_FilteringDone_clean_df)

# Convert the standardized vector into a matrix (826 rows, 102 columns)
stand_log_abundance_only_matrix <- matrix(stand_log_abundance_only, nrow = 826, ncol = 102)

# Optionally convert the standardized matrix into a data frame
stand_log_abundance_only_df <- as.data.frame(stand_log_abundance_only_matrix)

# Check the structure of the standardized dataset
str(stand_log_abundance_only_df)

# Scale the data by row to ensure each row has a mean of 0 and standard deviation of 1
abundance_FilteringDone_clean_scaled <- t(scale(t(abundance_FilteringDone_clean_df)))  # Row-wise scaling
stand_log_abundance_only_scaled <- t(scale(t(stand_log_abundance_only_df)))  # Row-wise scaling for standardized data

# Load the required library for heatmap visualization
library(pheatmap)

# Create a heatmap for the original data (before log transformation and standardization)
pheatmap(abundance_FilteringDone_clean_scaled,
         main = "Heatmap (Before Log Transformation and Standardisation)",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         scale = "none",  # No further scaling, since the data is already scaled row-wise
         show_rownames = FALSE,
         show_colnames = FALSE)

# Create a heatmap for the standardized data (after log transformation and standardization)
pheatmap(stand_log_abundance_only_scaled,
         main = "Heatmap (After Log Transformation and Standardisation)",
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         scale = "none",  # Data is already standardized
         show_rownames = FALSE,
         show_colnames = FALSE)

```
Transpose the abundance dataset
```{r}
# Check the structure of the original dataset
dim(abundance_FilteringDone_clean)  # Displays the number of rows and columns

# Convert log-transformed data back to a matrix (excluding metadata columns)
all_abundances_log_matrix <- matrix(
  stand_log_abundance_only,  # Log-transformed data vector
  nrow = nrow(abundance_FilteringDone_clean),  # Number of rows in the original dataset
  ncol = ncol(abundance_FilteringDone_clean) - 1  # Exclude metadata column (e.g., Protein names)
)

# Assign row and column names to the matrix
rownames(all_abundances_log_matrix) <- rownames(abundance_FilteringDone_clean)
colnames(all_abundances_log_matrix) <- colnames(abundance_FilteringDone_clean)[-1]  # Exclude metadata columns

# Verify the new structure
print(dim(all_abundances_log_matrix))
print(head(all_abundances_log_matrix))

# Transpose the matrix
abundance_data_transposed <- t(all_abundances_log_matrix)

# Convert the transposed matrix to a data frame
abundance_data_transposed_df <- as.data.frame(abundance_data_transposed)

# Add original row names as column names (Protein IDs)
colnames(abundance_data_transposed_df) <- rownames(all_abundances_log_matrix)
rownames(abundance_data_transposed_df) <- colnames(all_abundances_log_matrix)

# Add Protein IDs as column names and Sample IDs as row names
protein_ID <- abundance_FilteringDone_clean$Protein.ID  # Extract Protein IDs
colnames(abundance_data_transposed_df) <- protein_ID  # Set Protein IDs as column names

# Add an 'ID' column for the row names and rearrange
abundance_data_transposed_df$ID <- rownames(abundance_data_transposed_df)

# Move the 'ID' column to the first position
abundance_data_transposed_df <- abundance_data_transposed_df[, c("ID", setdiff(names(abundance_data_transposed_df), "ID"))]

# Remove original row names as they are now in the 'ID' column
rownames(abundance_data_transposed_df) <- NULL

# View the final transposed dataset with Protein IDs as column names and Sample IDs as row names
print("Transposed abundance dataset with Protein IDs as column names:")
head(abundance_data_transposed_df)

```
Merge the abundance dataset with the metadata dataset
```{r}
# Check column names in both datasets to ensure they have a common "ID" column
print("Metadata column names:")
print(colnames(metadata))

print("Abundance data column names:")
print(colnames(abundance_data_transposed_df))

# Merge the datasets by the "ID" column (Ensure all records are included from both datasets)
merged_data <- merge(metadata, abundance_data_transposed_df, by = "ID", all = TRUE)

# Preview the merged dataset
print("Merged dataset preview:")
head(merged_data)

# Save the merged dataset to a CSV file
write.csv(merged_data, "merged_data.csv", row.names = FALSE)

```
First hypothesis: What are the proteomic differences among HCC subtypes?

Anova test
```{r}
# Prepare a dataframe to store ANOVA results
results_anova <- data.frame(
  Protein_ID = colnames(merged_data)[22:ncol(merged_data)],  # Get protein names from columns 22 onwards
  P_Value = NA,  # Placeholder for p-values
  stringsAsFactors = FALSE
)

# Loop through each protein column and run the ANOVA test
for (i in 22:ncol(merged_data)) {  
  model <- aov(merged_data[, i] ~ merged_data$Proteomic_Subtype)  # Run ANOVA for each protein
  results_anova$P_Value[i - 21] <- summary(model)[[1]][["Pr(>F)"]][1]  # Extract p-value
}

# Apply Bonferroni correction to p-values
results_anova$Adj_P_Value_Bonferroni <- p.adjust(results_anova$P_Value, method = "bonferroni")

# Select significant proteins (p-value < 0.05 after Bonferroni correction)
significant_proteins_anova <- subset(results_anova, Adj_P_Value_Bonferroni < 0.05)

# Get the list of significant protein IDs
significant_proteins_list_anova <- significant_proteins_anova$Protein_ID

# Output the list of significant proteins and the count
cat("Significant proteins after Bonferroni correction:\n")
print(significant_proteins_list_anova)

# Count the number of significant proteins
aantal_significante_eiwitten <- nrow(significant_proteins_anova)

# Print the number of significant proteins
cat("Number of significant proteins after Bonferroni correction:", aantal_significante_eiwitten, "\n")

```
Tukey HSD post-hoc test
```{r}
# Dataframe to store results of post-hoc analyses
post_hoc_results <- data.frame(Protein_ID = character(),
                               Group_Comparison = character(), 
                               P_Value = numeric(),
                               Adj_P_Value = numeric(),
                               Lower_CI = numeric(),
                               Upper_CI = numeric(),
                               stringsAsFactors = FALSE)

# Loop through only significant proteins from the ANOVA test
for (protein in significant_proteins_list_anova) {
  
  # Find the index of the protein in the dataset
  protein_index <- which(colnames(merged_data) == protein)
  
  # Perform the ANOVA model for the protein
  model <- aov(merged_data[, protein_index] ~ merged_data$Proteomic_Subtype)
  
  # Perform Tukey's HSD post-hoc test to compare the subtypes
  tukey_result <- TukeyHSD(model)
  
  # Extract results from Tukey's test
  comparisons <- rownames(tukey_result$`merged_data$Proteomic_Subtype`)
  p_values <- tukey_result$`merged_data$Proteomic_Subtype`[, "p adj"]
  lower_ci <- tukey_result$`merged_data$Proteomic_Subtype`[, "lwr"]
  upper_ci <- tukey_result$`merged_data$Proteomic_Subtype`[, "upr"]
  
  # Add the results to the results dataframe
  for (i in 1:length(comparisons)) {
    post_hoc_results <- rbind(post_hoc_results, 
                              data.frame(Protein_ID = protein,
                                         Group_Comparison = comparisons[i],
                                         P_Value = p_values[i],
                                         Adj_P_Value = p.adjust(p_values[i], method = "bonferroni"), # Bonferroni correction for multiple tests
                                         Lower_CI = lower_ci[i],
                                         Upper_CI = upper_ci[i],
                                         stringsAsFactors = FALSE))
  }
}

# Select significant results (e.g., adjusted p-value < 0.05) and ensure confidence intervals do not include 0
significant_post_hoc_results <- post_hoc_results %>%
  filter(Adj_P_Value < 0.05 & !(Lower_CI <= 0 & Upper_CI >= 0))

# Overview of significant comparisons after Tukey's post-hoc test
print(significant_post_hoc_results)

```
Select non-tumor tissue columns
```{r}
# Select non-tumor tissue columns by selecting columns that end with "P"
abundances_P <- abundances %>%
  select(ends_with("P"))

# Remove the "P" suffix from column names
colnames(abundances_P) <- sub("P$", "", colnames(abundances_P))

# View the first few rows of the modified dataset
head(abundances_P)

# Check the updated column names to ensure "P" has been removed
cat("Updated column names in non-tumor tissue data:\n")
print(colnames(abundances_P))

```
Remove missing values in non-tumor tissue data
```{r}

# Convert character columns to numeric, if necessary
abundances_P <- abundances_P %>%
  mutate(across(where(is.character), ~ as.numeric(.)))

# Replace NA values with the minimum value in each row
abundances_P_Filtered <- abundances_P %>%
  rowwise() %>%
  mutate(across(everything(), ~ ifelse(is.na(.), min(c_across(everything()), na.rm = TRUE), .))) %>%
  ungroup()

# Check if there are any NAs left in the dataset
na_count <- sum(is.na(abundances_P_Filtered))
cat("There are", na_count, "NAs left in the dataset.\n")

# Identify rows with any remaining NAs
rows_with_na <- which(apply(abundances_P_Filtered, 1, function(x) any(is.na(x))))
cat("Rows with NAs:", rows_with_na, "\n")

# View the rows with NAs
head(abundances_P_Filtered[rows_with_na, ])
```
Log transformation on non-tumor tissue data
```{r}
# Transform the data using log1p
abundances_P_Filtered_log <- log1p(abundances_P_Filtered)
```
Standardisation on non-tumor tissue data
```{r}
# Standardize the log-transformed abundance data (mean = 0, sd = 1)
abundances_P_stand <- scale(abundances_P_Filtered_log)

# View the standardized data (first few rows)
head(abundances_P_stand)

# Check if the standardization was correctly performed by verifying the mean and standard deviation of each column
column_means <- apply(abundances_P_stand, 2, mean)  # Mean of each column (should be close to 0)
column_sds <- apply(abundances_P_stand, 2, sd)      # Standard deviation of each column (should be close to 1)

# Print the results to check the mean and standard deviation
cat("Means of each column (should be close to 0):\n")
print(column_means)

cat("Standard deviations of each column (should be close to 1):\n")
print(column_sds)

```
Transpose non-tumor tissue dataset
```{r}

# Transpose the standardized non-tumor tissue data
abundances_P_stand_transposed <- t(abundances_P_stand)

# Convert the transposed matrix back into a data frame
abundances_P_stand_transposed <- as.data.frame(abundances_P_stand_transposed)

# Set the column names based on the Protein.ID from the original dataset
colnames(abundances_P_stand_transposed) <- abundances$ID

# View the first few rows of the transposed dataset
head(abundances_P_stand_transposed)

# Add row names as a new column 'ID' using rownames_to_column function
abundances_P_stand_transposed <- abundances_P_stand_transposed %>%
  tibble::rownames_to_column(var = "ID")  # Adding row names as 'ID' column

# Move the 'ID' column to the first position (if necessary, to ensure correct order)
abundances_P_stand_transposed <- abundances_P_stand_transposed[, c("ID", setdiff(names(abundances_P_stand_transposed), "ID"))]

# Remove the row names (since they are now stored as a column)
rownames(abundances_P_stand_transposed) <- NULL

# View the result to ensure everything is correct
head(abundances_P_stand_transposed)

```
Top 10 proteins (Anova)
```{r}

# Make a dataframe to store top 10 results per subtype
top10_per_subtype_anova <- data.frame(
  Subtype = character(),
  Protein_ID = character(),
  Mean_Tumor = numeric(),
  Mean_Control = numeric(),
  Abs_Difference = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each subtype in the dataset
for (subtype in unique(merged_data$Proteomic_Subtype)) {
  
  # Filter tumor samples for the current subtype
  tumor_data <- merged_data[merged_data$Proteomic_Subtype == subtype, ]
  
  # Extract control samples for the same proteins from the control dataset
  control_data <- abundances_P_stand_transposed[, colnames(tumor_data[, 22:ncol(merged_data)])]
  
  # Iterate through significant proteins identified in post-hoc analysis
  significant_proteins <- unique(significant_post_hoc_results$Protein_ID)
  
  # Ensure only proteins available in tumor_data are analyzed
  available_proteins <- intersect(significant_proteins, colnames(tumor_data))
  
  # Process each protein
  for (protein in available_proteins) {
    # Calculate the mean abundance for tumor samples
    tumor_mean <- mean(tumor_data[[protein]], na.rm = TRUE)
    
    # Calculate the mean abundance for control samples
    control_mean <- mean(control_data[[protein]], na.rm = TRUE)
    
    # Calculate the absolute difference
    abs_difference <- abs(tumor_mean - control_mean)
    
    # Append the results to the dataframe
    top10_per_subtype_anova <- rbind(
      top10_per_subtype_anova,
      data.frame(
        Subtype = subtype,
        Protein_ID = protein,
        Mean_Tumor = tumor_mean,
        Mean_Control = control_mean,
        Abs_Difference = abs_difference,
        stringsAsFactors = FALSE
      )
    )
  }
}

# Select the top 10 proteins per subtype based on the absolute difference
top10_per_subtype_anova <- top10_per_subtype_anova %>%
  group_by(Subtype) %>%
  arrange(Subtype, desc(Abs_Difference)) %>%  # Sort by descending Abs_Difference within each Subtype
  slice_head(n = 10) %>%  # Retain the top 10 rows per Subtype
  ungroup()  # Remove grouping

# Display the final result
print(top10_per_subtype_anova)

```
Visualisation top 10 proteins (Anova)
```{r}

# Visualisation of the top 10 proteins per subtype with the largest difference in abundance
ggplot(top10_per_subtype_anova, aes(x = reorder(Protein_ID, -Abs_Difference), y = Abs_Difference, fill = Subtype)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Top 10 Proteins per Subtype (ANOVA)", 
       x = "Protein ID", 
       y = "Difference in Abundance (Tumor vs. Non-Tumor)") +  
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  
        axis.text.y = element_text(size = 8),  
        legend.position = "top", 
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) +  
  scale_fill_manual(values = custom_colors)  

```
Kruskal Wallis test
```{r}

# Create a dataframe to store results of the Kruskal-Wallis test
results_kruskal <- data.frame(
  Protein_ID = colnames(merged_data[, 22:ncol(merged_data)]),  # Extract protein column names
  P_Value = NA,  # Placeholder for p-values
  stringsAsFactors = FALSE
)

# Perform Kruskal-Wallis test for each protein and save the p-value
for (i in 22:ncol(merged_data)) {  # Loop through protein columns starting from column 22
  model <- kruskal.test(merged_data[, i] ~ merged_data$Proteomic_Subtype)  # Perform Kruskal-Wallis test
  results_kruskal$P_Value[i - 21] <- model$p.value  # Save the p-value in the results dataframe
}

# Apply Bonferroni correction to adjust the p-values for multiple comparisons
results_kruskal$Adj_P_Value_Bonferroni <- p.adjust(results_kruskal$P_Value, method = "bonferroni")

# Select significant proteins (adjusted p-value < 0.05 after Bonferroni correction)
significant_proteins_kruskal <- results_kruskal[results_kruskal$Adj_P_Value_Bonferroni < 0.05, ]

# Extract the list of significant protein IDs
significant_proteins_list_kruskal <- significant_proteins_kruskal$Protein_ID

# Count the number of significant proteins
num_significant_kruskal <- nrow(significant_proteins_kruskal)

# Print the count of significant proteins
cat("Number of significant proteins after Kruskal-Wallis test and Bonferroni correction:", num_significant_kruskal, "\n")

# View the list of significant proteins 
print(significant_proteins_list_kruskal)

```
Post hoc test: Dunn test
```{r}
# Load the necessary package for Dunn's test
library(dunn.test)

# Initialize a dataframe to store results from the Dunn's post-hoc test
post_hoc_kruskal_results <- data.frame(
  Protein_ID = character(),       # Protein identifier
  Group_Comparison = character(), # Pairwise group comparisons
  P_Value = numeric(),            # Adjusted p-value for each comparison
  stringsAsFactors = FALSE
)

# Loop through each significant protein (from Kruskal-Wallis test) to perform Dunn's test
for (protein in significant_proteins_list_kruskal) {
  
  # Run the Kruskal-Wallis test for the protein to confirm significance
  model <- kruskal.test(merged_data[, protein] ~ merged_data$Proteomic_Subtype)
  
  # Perform Dunn's post-hoc test only if Kruskal-Wallis is significant (p < 0.05)
  if (model$p.value < 0.05) {
    
    # Perform Dunn's test with Bonferroni adjustment
    dunn_result <- dunn.test(
      merged_data[, protein], 
      merged_data$Proteomic_Subtype, 
      method = "bonferroni"
    )
    
    # Extract pairwise comparisons and adjusted p-values from the Dunn's test result
    comparisons <- dunn_result$comparisons
    p_values <- dunn_result$P.adjusted
    
    # Add the results for this protein to the results dataframe
    for (i in 1:length(comparisons)) {
      post_hoc_kruskal_results <- rbind(
        post_hoc_kruskal_results, 
        data.frame(
          Protein_ID = protein,
          Group_Comparison = comparisons[i],
          P_Value = p_values[i],
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

# Filter the results to include only significant pairwise comparisons (p < 0.05)
significant_post_hoc_kruskal <- post_hoc_kruskal_results[
  post_hoc_kruskal_results$P_Value < 0.05, 
]

# Display the significant pairwise comparisons from Dunn's post-hoc test
print(significant_post_hoc_kruskal)

```
Top 10 eiwitten na Kruskal Wallis
```{r}
# Dataframe to store the top 10 results per subtype after Kruskal-Wallis and post-hoc tests
top10_per_subtype_kruskal <- data.frame(
  Subtype = character(),       # Subtype name
  Protein_ID = character(),    # Protein ID
  Mean_Tumor = numeric(),      # Mean abundance in tumor samples
  Mean_Control = numeric(),    # Mean abundance in control samples
  Abs_Difference = numeric(),  # Absolute difference between tumor and control
  stringsAsFactors = FALSE
)

# Loop through each subtype in the dataset
for (subtype in unique(merged_data$Proteomic_Subtype)) {
  
  # Filter tumor samples for the current subtype
  tumor_data <- merged_data[merged_data$Proteomic_Subtype == subtype, ]
  
  # Match the columns of control data to those of tumor data (adjust for protein columns)
  control_data <- abundances_P_stand_transposed[, colnames(tumor_data[, 22:ncol(merged_data)])]
  
  # Iterate over significant proteins from Dunn's post-hoc test
  for (protein in unique(significant_post_hoc_kruskal$Protein_ID)) {
    
    # Ensure the protein exists in the current subtype's data
    if (protein %in% colnames(tumor_data)) {
      
      # Calculate the mean abundance for tumor samples
      tumor_mean <- mean(tumor_data[, protein], na.rm = TRUE)
      
      # Calculate the mean abundance for control samples
      control_mean <- mean(control_data[, protein], na.rm = TRUE)
      
      # Calculate the absolute difference between tumor and control means
      abs_difference <- abs(tumor_mean - control_mean)
      
      # Append the results to the output dataframe
      top10_per_subtype_kruskal <- rbind(
        top10_per_subtype_kruskal, 
        data.frame(
          Subtype = subtype,
          Protein_ID = protein,
          Mean_Tumor = tumor_mean,
          Mean_Control = control_mean,
          Abs_Difference = abs_difference,
          stringsAsFactors = FALSE
        )
      )
    }
  }
}

# Filter and sort the results to select the top 10 proteins per subtype
library(dplyr)
top10_per_subtype_kruskal <- top10_per_subtype_kruskal %>%
  group_by(Subtype) %>%                   # Group by subtype
  arrange(Subtype, desc(Abs_Difference)) %>% # Sort by absolute difference (descending)
  slice_head(n = 10) %>%                  # Select top 10 proteins for each subtype
  ungroup()                               # Remove grouping for clean output

# Display the results
print(top10_per_subtype_kruskal)

```
Visualisation na Kruskal Wallis
```{r}
# Visualisation of the top 10 proteins per subtype from Kruskal-Wallis results
ggplot(top10_per_subtype_kruskal, aes(x = reorder(Protein_ID, -Abs_Difference), y = Abs_Difference, fill = Subtype)) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Top 10 proteins per Subtype (Kruskal-Wallis)",
       x = "Protein ID",
       y = "Difference in Abundance (Tumor vs. Non-Tumor)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),  
        axis.text.y = element_text(size = 8),  
        legend.position = "top",  
        plot.title = element_text(hjust = 0.5, size = 14, face = "bold")) +  
  scale_fill_manual(values = custom_colors)  

```
Comparison Anova vs Kruskal Wallis
```{r}
# Extract Protein_IDs and corresponding ranks from ANOVA and Kruskal-Wallis
anova_proteins <- unique(top10_per_subtype_anova$Protein_ID)  # Proteins from ANOVA
kruskal_proteins <- unique(top10_per_subtype_kruskal$Protein_ID)  # Proteins from Kruskal-Wallis

# Create a comparison dataframe with all proteins
comparison_full <- data.frame(
  Protein_ID = union(anova_proteins, kruskal_proteins),  # Include all proteins from both tests
  ANOVA_Rank = match(union(anova_proteins, kruskal_proteins), anova_proteins),
  Kruskal_Wallis_Rank = match(union(anova_proteins, kruskal_proteins), kruskal_proteins)
)

# Add a column to indicate proteins common to both tests
comparison_full$Matched <- ifelse(!is.na(comparison_full$ANOVA_Rank) & !is.na(comparison_full$Kruskal_Wallis_Rank), "Yes", "No")

# Sort the dataframe for clarity
comparison_full <- comparison_full[order(comparison_full$Matched, comparison_full$ANOVA_Rank, comparison_full$Kruskal_Wallis_Rank), ]

# Print the comparison
cat("Comparison of Proteins from ANOVA and Kruskal-Wallis:\n")
print(comparison_full)

# Output proteins unique to ANOVA or Kruskal-Wallis
anova_only <- comparison_full[is.na(comparison_full$Kruskal_Wallis_Rank), "Protein_ID"]
kruskal_only <- comparison_full[is.na(comparison_full$ANOVA_Rank), "Protein_ID"]

if (length(anova_only) > 0) {
  cat("\nProteins found only in ANOVA results:\n")
  print(anova_only)
}

if (length(kruskal_only) > 0) {
  cat("\nProteins found only in Kruskal-Wallis results:\n")
  print(kruskal_only)
}

```
Visualisation of the comparison
```{r}
# Filter the proteins that appear in only one subtype from the top 10 of ANOVA
unique_proteins_anova <- top10_per_subtype_anova %>%
  group_by(Protein_ID) %>%
  summarise(Count = n(), Subtypes = paste(unique(Subtype), collapse = ", ")) %>%
  filter(Count == 1)  # Select proteins that are unique to one subtype

# Add subtype information
unique_proteins_anova <- top10_per_subtype_anova %>%
  filter(Protein_ID %in% unique_proteins_anova$Protein_ID)

# Filter the proteins that appear in only one subtype from the top 10 of Kruskal-Wallis
unique_proteins_kruskal <- top10_per_subtype_kruskal %>%
  group_by(Protein_ID) %>%
  summarise(Count = n(), Subtypes = paste(unique(Subtype), collapse = ", ")) %>%
  filter(Count == 1)  # Select proteins that are unique to one subtype

# Add subtype information
unique_proteins_kruskal <- top10_per_subtype_kruskal %>%
  filter(Protein_ID %in% unique_proteins_kruskal$Protein_ID)

# Combine the results into one overview
unique_proteins_combined <- list(
  ANOVA_Unique = unique_proteins_anova,
  Kruskal_Unique = unique_proteins_kruskal
)

# Print the results
cat("\nProteins unique to one subtype in ANOVA top 10:\n")
print(unique_proteins_combined$ANOVA_Unique)

cat("\nProteins unique to one subtype in Kruskal-Wallis top 10:\n")
print(unique_proteins_combined$Kruskal_Unique)

# Add a column to indicate which test the data came from
unique_proteins_anova$Source <- "ANOVA"
unique_proteins_kruskal$Source <- "Kruskal-Wallis"

# Combine the datasets for both tests (ANOVA and Kruskal-Wallis)
combined_data <- rbind(unique_proteins_anova, unique_proteins_kruskal)

# Create a list of all unique proteins from both tests
all_proteins <- unique(combined_data$Protein_ID)

# Use a cheerful color palette (e.g., 'Set3' in RColorBrewer)
protein_colors <- setNames(colorRampPalette(brewer.pal(12, "Set3"))(length(all_proteins)), all_proteins)

# Specific exception for certain proteins: enforce unique colors
protein_colors["P05089"] <- "#FF6347"  # Tomato red for ANOVA
protein_colors["P08319"] <- "#1E90FF"  # Dodger blue for Kruskal-Wallis

# Visualization for ANOVA test
ggplot(unique_proteins_anova, aes(x = Subtype, y = Abs_Difference, fill = Protein_ID)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proteins per Subtype - ANOVA Test",
    x = "Subtype",
    y = "Abs. Difference in Mean Expression"
  ) +
  scale_fill_manual(
    name = "Proteins",
    values = protein_colors
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Visualization for Kruskal-Wallis test
ggplot(unique_proteins_kruskal, aes(x = Subtype, y = Abs_Difference, fill = Protein_ID)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(
    title = "Proteins per Subtype - Kruskal-Wallis Test",
    x = "Subtype",
    y = "Abs. Difference in Mean Expression"
  ) +
  scale_fill_manual(
    name = "Proteins",
    values = protein_colors
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

Second hypothesis: Is APF level a marker for tumor diameter?

Metadata exploration
```{r}
# View the metadata
metadata

# Select only the relevant columns from the dataset (ID, AFP in ng/ml, and tumor diameter in cm)
metadata_filtered <- merged_data %>% select(ID, AFP_.ng.ml., Diameter_of_tumor_.cm.)

# Check the first few rows of the filtered data to ensure it contains the correct columns
print(metadata_filtered)

# View the structure of the filtered dataset to understand the data types and summary of the variables
str(metadata_filtered)

```
Convert Diameter_of_tumor_.cm. to numeric
```{r}

# Create a copy of the original dataset so the original can be used later
metadata_filtered_numeric <- metadata_filtered

# Check the class of the 'Diameter_of_tumor_.cm.' column to ensure it is read as a character
class(metadata_filtered_numeric$Diameter_of_tumor_.cm.)

# Convert the 'Diameter_of_tumor_.cm.' column to numeric type for analysis
metadata_filtered_numeric$Diameter_of_tumor_.cm. <- as.numeric(metadata_filtered_numeric$Diameter_of_tumor_.cm.)

# Check again to confirm if the conversion to numeric was successful
class(metadata_filtered_numeric$Diameter_of_tumor_.cm.)

```
Localise NAs
```{r}
# Check for NA values, which may have been introduced due to cases where there are two diameters recorded
# View rows with NA values in the 'Diameter_of_tumor_.cm.' column
na_rows <- metadata_filtered_numeric[is.na(metadata_filtered_numeric$Diameter_of_tumor_.cm.), ]

# Display rows with NA values
print(na_rows)

```
Split the 'Diameter_of_tumor_.cm.' column by the "+" sign and create new rows for each diameter
```{r}
# Split the 'Diameter_of_tumor_.cm.' column by the "+" sign and create new rows for each diameter
merged_data_expanded <- metadata_filtered %>%
  separate_rows(Diameter_of_tumor_.cm., sep = "\\+")

# Check the result: view the first few rows to ensure the split worked correctly
print(merged_data_expanded)

# Convert the 'Diameter_of_tumor_.cm.' column to numeric after the split
metadata_modified <- merged_data_expanded %>%
  mutate(Diameter_of_tumor_.cm. = as.numeric(Diameter_of_tumor_.cm.))

# Check the structure of the modified dataset to ensure the column is now numeric
str(metadata_modified)  

# Print the modified dataset to inspect the rows
print(metadata_modified) 

```
Outlier detection
```{r}
# Log transformation for AFP (AFP should be positive for log transformation)
metadata_modified$log_AFP <- log(metadata_modified$AFP_.ng.ml.)

# Boxplot for AFP, log AFP and Tumor Diameter
par(mfrow = c(1, 3))  # Set the layout to display three plots side by side

# Boxplot for AFP (Alpha-fetoprotein) levels
boxplot(metadata_modified$AFP_.ng.ml., 
        main = "Boxplot of AFP Levels", 
        ylab = "AFP (ng/ml)", 
        col = "lightblue", 
        border = "darkblue",  # Darker border for better contrast
        notch = FALSE,        # Remove notch to avoid warning
        horizontal = FALSE,   # Vertical boxplot for better readability
        outline = TRUE)       # Include outliers

# Boxplot for log(AFP)
boxplot(metadata_modified$log_AFP, 
        main = "Boxplot of log(AFP)", 
        ylab = "log(AFP)", 
        col = "lavender", 
        border = "purple4",   # Darker border for better contrast
        notch = FALSE,         # Remove notch to avoid warning
        horizontal = FALSE,    # Vertical boxplot for better readability
        outline = TRUE)        # Include outliers

# Boxplot for Tumor Diameter 
boxplot(metadata_modified$Diameter_of_tumor_.cm., 
        main = "Boxplot of Tumor Diameter",
        ylab = "Tumor Diameter (cm)", 
        col = "lightpink",     # Change the fill color to pink
        border = "darkred",    # Darker red border for contrast
        notch = FALSE,         # Remove notch to avoid warning
        horizontal = FALSE,    # Vertical boxplot for better readability
        outline = TRUE)        # Include outliers

# Function to calculate Z-scores and IQRs
calculate_outliers <- function(data_column) {
  z_scores <- scale(data_column)
  iqr_value <- IQR(data_column)
  Q1 <- quantile(data_column, 0.25)
  Q3 <- quantile(data_column, 0.75)
  lower <- Q1 - 1.5 * iqr_value
  upper <- Q3 + 1.5 * iqr_value
  
  list(z_scores = z_scores, lower = lower, upper = upper, Q1 = Q1, Q3 = Q3)
}

# Apply the outlier detection for AFP, log(AFP), and Tumordiameter
AFP_outliers <- calculate_outliers(metadata_modified$AFP_.ng.ml.)
log_AFP_outliers <- calculate_outliers(metadata_modified$log_AFP)
diameter_outliers <- calculate_outliers(metadata_modified$Diameter_of_tumor_.cm.)

# Detect outliers using Z-scores (threshold > 3 or < -3)
outliers_z_AFP <- which(abs(AFP_outliers$z_scores) > 3)
outliers_z_log_AFP <- which(abs(log_AFP_outliers$z_scores) > 3)
outliers_z_diameter <- which(abs(scale(metadata_modified$Diameter_of_tumor_.cm.)) > 3)

# Detect outliers using IQR method
outliers_IQR_AFP <- metadata_modified[metadata_modified$AFP_.ng.ml. < AFP_outliers$lower | metadata_modified$AFP_.ng.ml. > AFP_outliers$upper, ]
outliers_IQR_log_AFP <- metadata_modified[metadata_modified$log_AFP < log_AFP_outliers$lower | metadata_modified$log_AFP > log_AFP_outliers$upper, ]
outliers_IQR_diameter <- metadata_modified[metadata_modified$Diameter_of_tumor_.cm. < diameter_outliers$lower | metadata_modified$Diameter_of_tumor_.cm. > diameter_outliers$upper, ]

# Combine Z-score and IQR outliers for AFP and Tumordiameter
outliers_combined_AFP <- unique(c(outliers_z_AFP, rownames(outliers_IQR_AFP)))
outliers_combined_log_AFP <- unique(c(outliers_z_log_AFP, rownames(outliers_IQR_log_AFP)))
outliers_combined_diameter <- unique(c(outliers_z_diameter, rownames(outliers_IQR_diameter)))

# Print combined outliers
cat("Combined outliers for AFP (original):\n")
print(metadata_modified[outliers_combined_AFP, c("AFP_.ng.ml.", "Diameter_of_tumor_.cm.")])

cat("\nCombined outliers for log(AFP):\n")
print(metadata_modified[outliers_combined_log_AFP, c("log_AFP", "Diameter_of_tumor_.cm.")])

cat("\nCombined outliers for Tumordiameter:\n")
print(metadata_modified[outliers_combined_diameter, c("AFP_.ng.ml.", "Diameter_of_tumor_.cm.")])

```
Correlation
```{r}
# Remove outliers using combined outlier indices for AFP
metadata_no_outliers <- metadata_modified[!rownames(metadata_modified) %in% outliers_combined_AFP, ]

# Pearson and Spearman correlation for AFP and Tumor Diameter (with outliers)
correlation_AFP_diameter_outliers_pearson <- cor(metadata_modified$AFP_.ng.ml., metadata_modified$Diameter_of_tumor_.cm., method = "pearson")
correlation_AFP_diameter_outliers_spearman <- cor(metadata_modified$AFP_.ng.ml., metadata_modified$Diameter_of_tumor_.cm., method = "spearman")

# Pearson and Spearman correlation for AFP and Tumor Diameter (without outliers)
correlation_AFP_diameter_pearson <- cor(metadata_no_outliers$AFP_.ng.ml., metadata_no_outliers$Diameter_of_tumor_.cm., method = "pearson")
correlation_AFP_diameter_spearman <- cor(metadata_no_outliers$AFP_.ng.ml., metadata_no_outliers$Diameter_of_tumor_.cm., method = "spearman")

# Pearson and Spearman correlation for log(AFP) and Tumor Diameter (no outliers to remove here)
correlation_log_AFP_diameter_pearson <- cor(metadata_modified$log_AFP, metadata_modified$Diameter_of_tumor_.cm., method = "pearson")
correlation_log_AFP_diameter_spearman <- cor(metadata_modified$log_AFP, metadata_modified$Diameter_of_tumor_.cm., method = "spearman")

# Print correlations
cat("\nPearson correlation for AFP and Tumor Diameter (without outliers removed):\n", correlation_AFP_diameter_outliers_pearson)
cat("\nSpearman correlation for AFP and Tumor Diameter (without outliers removed):\n", correlation_AFP_diameter_outliers_spearman)

cat("\nPearson correlation for AFP and Tumor Diameter (with outliers removed):\n", correlation_AFP_diameter_pearson)
cat("\nSpearman correlation for AFP and Tumor Diameter (with outliers removed):\n", correlation_AFP_diameter_spearman)

cat("\nPearson correlation for log(AFP) and Tumor Diameter (log transformation):\n", correlation_log_AFP_diameter_pearson)
cat("\nSpearman correlation for log(AFP) and Tumor Diameter (log transformation):\n", correlation_log_AFP_diameter_spearman)

# Plot the relationship between AFP (NOT filtered for outliers) and Tumor Diameter
ggplot(metadata_modified, aes(x = AFP_.ng.ml., y = Diameter_of_tumor_.cm.)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = paste("(Not Filtered) AFP vs Tumor Diameter  (Pearson: ", round(correlation_AFP_diameter_outliers_pearson, 2), ")", 
                     sep = ""),
       x = "AFP (ng/ml)", y = "Tumor Diameter (cm)")

# Plot the relationship between AFP (filtered for outliers) and Tumor Diameter
ggplot(metadata_no_outliers, aes(x = AFP_.ng.ml., y = Diameter_of_tumor_.cm.)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = paste("(Filtered) AFP vs Tumor Diameter (Pearson: ", round(correlation_AFP_diameter_pearson, 2), ")", 
                     sep = ""),
       x = "AFP (ng/ml)", y = "Tumor Diameter (cm)")

# Plot the relationship between log(AFP) (no outliers removed) and Tumor Diameter
ggplot(metadata_modified, aes(x = log_AFP, y = Diameter_of_tumor_.cm.)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = paste("log(AFP) vs Tumor Diameter (Pearson: ", round(correlation_log_AFP_diameter_pearson, 2), ")", 
                     sep = ""),
       x = "log(AFP) (ng/ml)", y = "Tumor Diameter (cm)")

```
Regression analysis
```{r}
# Regression analysis for AFP and Tumordiameter (with outliers)
model_AFP <- lm(Diameter_of_tumor_.cm. ~ AFP_.ng.ml., data = metadata_modified)
cat("\nRegression analysis (AFP with outliers):\n")
summary(model_AFP)

# Residual plot for AFP and Tumordiameter (with outliers)
ggplot(metadata_modified, aes(x = AFP_.ng.ml., y = residuals(model_AFP))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Residuals for AFP vs Tumordiameter (with outliers)", x = "AFP (ng/ml)", y = "Residuals")

# Regression analysis for AFP and Tumordiameter (without outliers)
model_AFP_no_outliers <- lm(Diameter_of_tumor_.cm. ~ AFP_.ng.ml., data = metadata_no_outliers)
cat("\nRegression analysis (AFP without outliers):\n")
summary(model_AFP_no_outliers)

# Residual plot for AFP and Tumordiameter (without outliers)
ggplot(metadata_no_outliers, aes(x = AFP_.ng.ml., y = residuals(model_AFP_no_outliers))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Residuals for AFP vs Tumordiameter (without outliers)", x = "AFP (ng/ml)", y = "Residuals")

# Regression analysis for log(AFP) and Tumordiameter (no outlier removal for log(AFP) and Tumor Diameter)
model_log_AFP <- lm(Diameter_of_tumor_.cm. ~ log_AFP, data = metadata_modified) 
cat("\nRegression analysis (log(AFP) without outliers):\n")
summary(model_log_AFP)

# Residual plot for log(AFP) and Tumordiameter (no outliers removed for log(AFP))
ggplot(metadata_modified, aes(x = log_AFP, y = residuals(model_log_AFP))) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", col = "red") +
  labs(title = "Residuals for log(AFP) vs Tumordiameter (without outliers)", x = "log(AFP) (ng/ml)", y = "Residuals")

```